Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-15 article.Rmd
Adding citations and editing

This paper addresses two trends in the humanities: the rise of gender as an analytical tool and the proliferation of large, machine-readable datasets. It outlines a new method for algorithmically inferring the gender of a name from historical data in order to take into account how naming practices change over time and space. In contrast to existing methods of predicting gender, this method provides a higher level of accuracy and a precise statement of the likelihood that a name is a given gender. The article describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to demonstrate the method's utility by applying it to a case study in which we study gender in the history profession by inferring the gender of authors of history dissertations and of authors and reviewers in the *American Historical Review*.

@@

As historians and other humanities researchers grapple with larger and larger datasets, they are increasingly using algorithms not just to extract information from existing data but to infer additional information from the data that is available.^[CITE: find example – network analysis, maybe http://kieranhealy.org/blog/archives/2013/06/09/using-metadata-to-find-paul-revere/] Datasets of interest to historians---for example membership lists, company payrolls, military rosters, passenger bookings, records of correspondence, and lists of published works---might included records with the names of individual people. Some datasets might contain associated information like a person's address, age, or rank. Other datasets contain little more than a personal name. Given large but sparse datasets, researchers have to use their ingenuity to derive additional kinds of information. What kind of information can researchers infer from just a name? Most obviously first names often imply a person's gender. "Jane Fay" is almost certainly a woman, while her brother "John Fay" is almost certainly a man. A straightforward computer program can infer their respective genders with relative ease.

The problem arises from the fact that the link between gender and naming practices, like language itself, is not static. What about Jane and John Fay's sibling, Leslie? If Leslie was born in the past sixty years, chances are good that Leslie would be their sister. But if the three siblings were born in the early twentieth century, chances are good that Leslie would be their brother. That is because the gendered naming convention surrounding "Leslie" completely changed over the course of one hundred years. 92% of the babies born in the United States in 1900 who were named "Leslie" were male, while 96% of those identically named babies born in the year 2000 were female.

@@

Many other names as used in the United States, such as Madison, Morgan, Sydney, and Kendall, have also changed gender over the past century. Such changes usually follow set patterns. In general, names tend to shift from being predominantly male to being predominantly female; the shift rarely moves in the other direction. In general, names that transition from male to female also became increasingly popular over time. For example, the name "Madison" changed from male to female suddenly after 1985. Whereas before it had been an infrequently used male name, after 1985 it skyrocketed in popularity, and by the early 2000s it was one of the ten most frequently used names for baby girls. Broadly speaking, since the late twentieth century Americans girls are more likely than American boys to receive a "trendy" name, and once that name is associated with females it is unlikely to be given to male babies.^[The few exceptions of names which go from being female to male (such as "Jan") have a different causal mechanism. It seems likely that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change.]

@@

For researchers working on contemporary subjects the "Leslie" problem is not an especially pressing one. An education policy analyst studying the demographics of different urban school districts, for instance, can turn to a variety of tools that use current databases of names, and these tools will likely identify the gender of Leslie Fay correctly. For example, the [Genderize.io](http://genderize.io/) online service determines the gender of first names "from user profiles across major social networks." Checking the service for the name Leslie returns a prediction that the name is female, along with an estimate of what proportion of the uses among members of unnamed social media sites is female and male.^[In this example the Genderize.io service has been accessed through the Gender package described below.]

@@

Researchers studying historical subjects, on the other hand, need to take into account changes in naming practices. If that same researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, contemporary datasets would misidentify Leslie Fay's gender. The "Leslie" problem is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose characteristics do not match the probably unknowable characteristics of the groups that comprise social media datasets will also encounter this problem. As of 2012, the average American has a lifespan of nearly seventy-nine years–--more than enough time for naming practices to change quite dramatically.^[World Bank, *World Bank Open Data*, "Life expectancy at birth, total (years)," <http://data.worldbank.org/indicator/SP.DYN.LE00.IN> (accessed September 12, 2014).] Predicting gender from first names therefore requires a fundamentally historical method.

@@

>>>>

This paper addresses two trends in the humanities: the rise of gender as an analytical tool and the proliferation of large, machine-readable datasets. It outlines a new method for algorithmically inferring the gender of a name from historical data in order to take into account how naming practices change over time and space. In contrast to existing methods of predicting gender, this method provides a higher level of accuracy and a precise statement of the likelihood that a name is a given gender. The article describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to apply the method to a case study in which we study gender and gatekeeping in the history profession by inferring the gender of authors of history dissertations and of authors and reviewers in the *American Historical Review*.

@@

As humanities researchers grapple with larger and larger datasets they will increasingly be able to not just extract existing information from that data but to also infer additional information from it \[Goldstone and Underwood 2014]\. ^[The availability of large-scale datasets dealing with personal information raises important ethical issues for digital humanists, particularly in light of the revelations about domestic spying at the NSA brought forth by Edward Snowden in 2013. \[Weingart 2014]\ Kieran Healy has used network analysis to demonstrate how researchers can infer information from datasets while also touching on the troubling ethical issue of surveillance \[Kiely 2013]\.] Many of these datasets deal with lists of people: company payrolls, military rosters, passenger bookings, records of correspondence, or lists of published works. Some of these records might contain associated information such a person's address, age, or rank, while others contain little more than a list of personal names. Given large but sparse datasets, researchers have to use their ingenuity to derive additional kinds of information. What can researchers infer from just a name? On a basic level, first names often imply a person's gender. "Jane Fay" is almost certainly a woman, while her brother "John Fay" is almost certainly a man. A straightforward computer program can infer their respective genders with relative ease.

The problem with inferring gender from names is that the link between gender and naming practices, like language itself, is not static. What about Jane and John Fay's sibling, Leslie? If Leslie was born in the past sixty years, chances are good that Leslie would be their sister. But if the three siblings were born in the early twentieth century, chances are good that Leslie would be their brother. That is because the gendered naming convention surrounding "Leslie reversed course completely over the course of one hundred years. 92% of the babies born in the United States in 1900 who were named "Leslie" were male, while 96% of those identically named babies born in the year 2000 were female.

@@

Other names used in the United States such as Madison, Morgan, Sydney, and Kendall, have also "flipped" their gender over the past century. Such changes usually follow set patterns. In general, names tend to shift from being predominantly male to being predominantly female; the shift rarely moves in the other direction. In general, names that transition from male to female also became increasingly popular over time. For example, the name "Madison" changed from male to female quite suddenly after 1985. Whereas before it had been an infrequently used male name, after 1985 it skyrocketed in popularity and by the early 2000s was one of the ten most frequently used names for baby girls. Broadly speaking, in the late twentieth century Americans girls became increasingly more likely than American boys to receive a "trendy" name, and once that name was associated with women it was unlikely to be given to male babies. These names present a moving target for researchers hoping to use them to infer additional information about the people under study. ^[The few exceptions of names which flip from being female to male, such as "Jan", have a different causal mechanism. It seems likely that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change.]

@@

For those working on contemporary subjects the "Leslie" problem is not an especially pressing one. An education policy analyst studying the demographics of different urban school districts, for instance, can turn to a variety of tools that use current databases of names, and these tools will likely identify the gender of Leslie Fay correctly. One online service, [Genderize.io](http://genderize.io/), determines the gender of first names "from user profiles across major social networks." Checking the service for the name Leslie returns a prediction that the name is female, along with an estimate of male/female proportions of the uses among members of unnamed social media sites.^[In this example the Genderize.io service has been accessed through the Gender package described below.]

@@

But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same education researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, contemporary tools like Genderize.io would misidentify Leslie Fay's gender in historical datasets. The "Leslie" problem is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match those of a contemporary tool like Genderize.io will also encounter this problem. As of 2012, the average American has a lifespan of nearly seventy-nine years–--more than enough time for naming practices to change quite dramatically between their birth and death.^[World Bank, *World Bank Open Data*, "Life expectancy at birth, total (years)," <http://data.worldbank.org/indicator/SP.DYN.LE00.IN> (accessed September 12, 2014).] Predicting gender from first names therefore requires a fundamentally historical method.

@@

### References

Goldstone and Underwood 2014. Goldstone, Andrew, and Ted Underwood. "The Quiet Transformations of Literary Studies: What Thirteen Thousand Scholars Could Tell Us." *New Literary History* (Forthcoming 2014).

Healy 2013. Healy, Kieran. "Using Metadata to Find Paul Revere." *kieranhealy.org*. 9 June 2013, http://kieranhealy.org/blog/archives/2013/06/09/using-metadata-to-find-paul-revere/

Weingart 2014. Weingart, Scott. "The Moral Role of DH in a Data-Driven World." *the scottbot irregular*. 9 September 2014, http://www.scottbot.net/HIAL/?p=40944
